---
title: "dev"
author: "Sophie Castel"
date: "1/27/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<<<<<<< HEAD
#########################################################################


## JOSS 11 (Closed)
=======
## JOSS 11
>>>>>>> f435d1f... Updated issue log

```{r}
# loading uncompressed data files

load(file = "/home/sophie/subMaldi/data/After1.rda")
load(file = "/home/sophie/subMaldi/data/After2.rda")
load(file = "/home/sophie/subMaldi/data/Before1.rda")
load(file = "/home/sophie/subMaldi/data/Before2.rda")
load(file = "/home/sophie/subMaldi/data/Blank1.rda")
load(file = "/home/sophie/subMaldi/data/Blank2.rda")
load(file = "/home/sophie/subMaldi/data/bsline.rda")
load(file = "/home/sophie/subMaldi/data/Master.rda")
load(file = "/home/sophie/subMaldi/data/Master2.rda")

# test compress, comparing methods
save(Master, file = "/home/sophie/subMaldi/data/compressed/Master_T.rda", compress = TRUE)
save(Master, file = "/home/sophie/subMaldi/data/compressed/Master_xz.rda", compress = "xz")

file.info("/home/sophie/subMaldi/data/compressed/Master_T.rda")
file.info("/home/sophie/subMaldi/data/compressed/Master_xz.rda")

# opening test compress
load("/home/sophie/subMaldi/data/compressed/Master_xz.rda")

# saving else compressed files
save(After1, file = "/home/sophie/subMaldi/data/compressed/After1.rda", compress = "xz")
save(After2, file = "/home/sophie/subMaldi/data/compressed/After2.rda", compress = "xz")
save(Before1, file = "/home/sophie/subMaldi/data/compressed/Before1.rda", compress = "xz")
save(Before2, file = "/home/sophie/subMaldi/data/compressed/Before2.rda", compress = "xz")
save(Blank1, file = "/home/sophie/subMaldi/data/compressed/Blank1.rda", compress = "xz")
save(Blank2, file = "/home/sophie/subMaldi/data/compressed/Blank2.rda", compress = "xz")
save(bsline, file = "/home/sophie/subMaldi/data/compressed/bsline.rda", compress = "xz")
save(Master, file = "/home/sophie/subMaldi/data/compressed/Master.rda", compress = "xz")
save(Master2, file = "/home/sophie/subMaldi/data/compressed/Master2.rda", compress = "xz")

```

<<<<<<< HEAD

#########################################################################





## JOSS 5 (Closed)
Wrote biocViews: into DESCRIPTION install_github() can search BioConductor for MassSpecWavelet




#########################################################################




## JOSS 12 - generalize and fix redundancies in code (Closed)

## avgSpectra.R (done)
=======
## JOSS 5 
Wrote biocViews: into DESCRIPTION install_github() can search BioConductor for MassSpecWavelet

## JOSS 12 - generalize and fix redundancies in code

<<<<<<< HEAD
# avgSpectra.R
>>>>>>> f435d1f... Updated issue log
=======
# avgSpectra.R (done)
>>>>>>> d7c9654... Updated issue log

```{r}
avgSpectra <- function(dat, method = "mean", spectra_cols){
  
  # --------------
  # LOGICAL CHECKS
  # --------------
  
  if(length(spectra_cols) < 2){
    stop("Only one spectrum input. Please enter two spectra for averaging.")
  }
  
  if(!all(spectra_cols %in% colnames(dat))){
    logic <- which(!(spectra_cols %in% colnames(dat)))
    stop(c("Columns '",paste0(as.character(spectra_cols[logic]), sep = "', "), " not found in specified dataframe."))
  }
  
  if(method == "sum"){
    .avg_sum(dat = dat, spectra_cols) } 
  else if(method == "mean"){
    .avg_mean(dat = dat, spectra_cols) }
}

# --------------
# METHOD = SUM
# --------------

.avg_sum <- function(dat, spectra_cols){
  
  mz <- dat$full_mz
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  dat <- cbind(mz, dat, Sum = apply(i, 1, sum, na.rm = TRUE))
  
  return(dat)
}

# --------------
# METHOD = MEAN
# --------------

.avg_mean <- function(dat, spectra_cols){
  
  mz <- dat$full_mz
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  dat <- cbind(mz, dat, Average = apply(i, 1, mean, na.rm = TRUE))
  
  return(dat)
}

```

<<<<<<< HEAD
<<<<<<< HEAD
## find_max.R (done)
=======
# find_max.R
>>>>>>> f435d1f... Updated issue log
=======
# find_max.R (done)
>>>>>>> d7c9654... Updated issue log

```{r}
# -----------------
# FIND PEAK MAXIMA
# -----------------

# Find peak maximum and associated m/z of spectra
find_max <- function (dat, mass_dat, spectra_cols){
  
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  rownames(i) <- mz
  
  max_i <- apply(i, 2, max)
  which_max <- apply(i, 2, which.max)
  max_mz <- mz[which_max]
  
  max_spec <- cbind(max_i, max_mz)
  colnames(max_spec) <- c("Intensity (Max)", "Mass")
  
  return(max_spec)
}

# -----------------------------------------------------------------------

# ------------------------
# FIND PEAK MAXIMA OF SET
# ------------------------

find_max_set <- function(dat, mass_dat, spectra_cols){

  # --------------
  # LOGICAL CHECKS
  # --------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat))
  )
  
  
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  rownames(i) <- mz
  
  which_max <- which(i == max(i), arr.ind = TRUE)
  
  max_spec <- data.frame(Intensity = as.numeric(i[which_max[1], which_max[2]]), Mass = as.numeric(rownames(which_max)))
  
  colnames(max_spec) <- c("Intensity (Max)", "Mass")
  rownames(max_spec) <- colnames(i)[which_max[2]]
  
  return(max_spec)
  
}
```

<<<<<<< HEAD
## norm_custimp.R (done)

```{r}
# --------------------------------------
# METHOD: CUSTOM M/Z VALUE, NOT PRECISE
# --------------------------------------

norm_custimp <- function(dat, mass_dat, norm_mz, spectra_cols, showHI = FALSE){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat))
  )
  
  if(is.null(norm_mz)){
    stop('Please select a m/z value. norm_mz is NULL.')
  }   
  
  
  mz <- dat[[mass_dat]]
  
  logic_i <- which(startsWith(as.character(mz), as.character(norm_mz)))
  
  if(length(logic_i) == 0){
    stop(c("The selected value norm_mz = ", norm_mz, " not found in '", mass_dat,"'. Please try another value."))
  }
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  i_cust <- i[logic_i,]
  
  if(nrow(i_cust) > 1){
    stop("More than one peak per spectra for the given norm_mz value. Please be more precise.")    
  }
  
  i_cust <- as.numeric(i_cust)
  
  logic <- i_cust == 0 
  
  if(any(logic)){
    stop(c("The selected maximum intensity is 0 in spectra labeled: ", paste0(colnames(i)[logic], sep = " ")))
  }
  
  i_norm <- t(t( i - min(i) )* ( i_cust - min(i) )^-1) # (i-min(i))/(i_cust-min(i)) Matrix multiplication
  
  if(!showHI){
    i_norm[i_norm > 1] <- 1
  }
  
  dat <- data.frame(mz, i_norm)
  colnames(dat) <- c("full_mz", spectra_cols)
  
  return(dat)
}

```

## norm_custom.R (done)

```{r}

# -------------------------
# METHOD: CUSTOM M/Z VALUE
# -------------------------

.norm_custom <- function(y, custom_y){
  return((y - min(y)) / (custom_y - min(y))) # single value
}

.normMethod_custom <- function(dat, mass_dat, norm_mz, spectra_cols, showHI = FALSE){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
    
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat))
  )
    

    
  if(is.null(norm_mz)){
    stop('Please select a m/z value. norm_mz is NULL.')
  } 
  
    
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
    
  i_cust <- as.numeric(i[which(mz == norm_mz),])
  
  if(all(is.na(i_cust))){
    stop(c("The selected value norm_mz = ", norm_mz, " not found in '", mass_dat,"'. Please try another value."))
  }
  
  logic <- i_cust == 0 
    
  if(any(logic)){
    stop(c("The selected maximum intensity is 0 in spectra labeled: ", paste0(colnames(i[logic]), sep = " ")))
  }
    
  i_norm <- t(t( i - min(i) )* ( i_cust - min(i) )^-1) # (i-min(i))/(i_cust-min(i)) Matrix multiplication
    
  if(!showHI){
    i_norm[i_norm > 1] <- 1
  }
    
  dat <- data.frame(mz, i_norm)
  colnames(dat) <- c("full_mz", spectra_cols)
    
  return(dat)
    
}

```

## norm_max.R (done)
=======
# normMethod_custimp.R

```{r}
```

# normMethod_custom.R

```{r}
```

<<<<<<< HEAD
# normMethod_max.R
>>>>>>> f435d1f... Updated issue log
=======
# normMethod_max.R (done)
>>>>>>> d7c9654... Updated issue log

```{r}
# -----------------------
# NORMALIZATION FUNCTION
# -----------------------

# Function rescales intensity to 0,1
.normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}

# ------------------------------
# METHOD: MAX. OF EACH SPECTRUM
# ------------------------------

<<<<<<< HEAD
norm_max <- function(dat, mass_dat, spectra_cols){
=======
.normMethod_max <- function(dat, mass_dat, spectra_cols){
>>>>>>> f435d1f... Updated issue log

  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat))
  )
  
  
  
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  i_n <- apply(i, 2, FUN = .normalize)
  dat <- cbind(mz, i_n)
  colnames(dat) <- c("full_mz", spectra_cols)
  
  return(dat)
}
  
# ---------------------
# METHOD: MAX. OF SET
# ---------------------

# Normalization function specific to custom max_y
.normalize_set <- function(y, max_y){
  return((y - min(y)) / (max_y - min(y)))
}

.normMethod_max_set <- function(dat, mass_dat, spectra_cols){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat))
  )

  if(length(spectra_cols) < 2){
    stop("Only one spectrum input. Please enter two spectra.")
  }
  
  
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  which_max <- which(i == max(i), arr.ind = TRUE)
  max_i <- i[which_max[1], which_max[2]]
  
  i_n <- apply(i, 2, FUN = function(x) {.normalize_set(y = x, max_y = max_i)} )
  dat <- cbind(mz, i_n)
  colnames(dat) <- c("full_mz", spectra_cols)
  
  return(dat)
}

```

<<<<<<< HEAD
<<<<<<< HEAD
## norm_median.R (done)
=======
# normMethod_median.R
>>>>>>> f435d1f... Updated issue log
=======
# normMethod_median.R (done)
>>>>>>> d7c9654... Updated issue log

```{r}

# ---------------
# METHOD: MEDIAN
# ---------------

<<<<<<< HEAD
norm_med<- function(dat, mass_dat, spectra_cols){
=======
.normMethod_med<- function(dat, mass_dat, spectra_cols){
>>>>>>> f435d1f... Updated issue log

  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat))
  )

  if(length(spectra_cols) < 2){
    stop("Only one spectrum input. Please enter two spectra.")
  }
  
  
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  i[i == 0 ] <- NA
  
  i_med <- apply(i,2, FUN = median, na.rm = TRUE)
  max_med <- max(i_med)
  
  for(j in 1:length(spectra_cols)){
    
    # multiplicative factors
    factors <- c()
    
      if(i_med[j] == max_med){
        
        other <- i_med[-j]
        other_index <- which(i_med %in% other)
          
        factors <- rep(1,length(spectra_cols))
        
        factors[other_index] = i_med[j]/i_med[other_index]
        
      }
    }
    
  #i_mult <- i*factors
  
  i_mult <- t(t(i)*factors)
  
  dat <- cbind(mz, i_mult)
  colnames(dat) <- c("full_mz", spectra_cols)
  
  return(dat)
  
}
```

<<<<<<< HEAD
<<<<<<< HEAD
## norm_quantile.R (done)

```{r}
# --------------------------------------------------------------------------------------------
# Date: January 28, 2021
# Author: Kristen Yeh, Sophie Castel
# Title: Normalization Method - Quantile
# --------------------------------------------------------------------------------------------


# Quantile normalization consists of two steps:
    # 1. Create a mapping between ranks and values. For rank 1, find the n values,
       # one per spectrum, that are the smallest value on the spectrum, and save their
       # averages. Similarly for rank 2 and the second smallest values, and on up to
       # the n largest values, one per spectrum.
    # 2. For each spectrum, replace the actual values with these averages.

# Basically, index each intensity value in each spectrum then sort
# by intensity (value). Probably need 2 truncate. Average spectrum for the sorted spectra
# is calculated. The averages are then inserted into the sorted spectra.
# The spectra are then reverted to their unsorted order using the index.


norm_quantile <- function(dat, mass_dat, spectra_cols){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat)),
    is.numeric(lower),
    is.numeric(upper)
  )    
  
  if(length(spectra_cols) < 2){
    stop("Only one spectrum input. Please enter two spectra for comparison.")
    }
 
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  i <- data.frame(mz, i)
  colnames(i) <- c("full_mz", spectra_cols)
  
  i_melt <- melt(i, id.vars = "full_mz")
  colnames(i_melt) <- c("full_mz","Spectrum","Intensity")
  
  i_melt <- dplyr::filter(i_melt, Intensity != 0)
  
  i_melt <- i_melt %>% group_by(Spectrum)  %>% mutate(id = row_number())
  ordr <- i_melt[order(i_melt$Intensity, decreasing = TRUE),]
  
  r <- rep(NA, length(spectra_cols))
  
  for(j in 1:length(spectra_cols)){
    r[j] <- max(as.numeric(i_melt$id[which(i_melt$Spectrum == spectra_cols[j])]))
  }
  
  r_min <- min(r)
  
  logic <-which(r == r_min)
  
  trunc <- ordr[(which(ordr$id <= r[logic])),]
  
  i_trunc <- data.frame(matrix(ncol = length(unique(trunc$Spectrum)), nrow = max(trunc$id)))
  
  for(j in 1:length(unique(trunc$Spectrum))){
    
    i_trunc[j] <- trunc$Intensity[which(trunc$Spectrum == spectra_cols[j])]
    colnames(i_trunc)[j] <- spectra_cols[j]
    
  }
  
  i_trunc <- data.frame(i_trunc, avg = apply(i_trunc, 1, FUN = mean, na.rm = TRUE))
  
  for(j in 1:length(unique(trunc$Spectrum))){
    trunc$Intensity[which(trunc$Spectrum == colnames(i_trunc)[j])] <- i_trunc$avg
  }
  
  dat <- trunc[order(trunc[mass_dat], decreasing = FALSE), ] %>% spread(Spectrum, Intensity)
  
  dat[is.na(dat)] <- 0
  dat <- dat[-2]
  return(as.data.frame(dat))
}

```

## norm_RMS.R (done)

```{r}

# -------------------------
# METHOD: ROOT MEAN SQUARE
# -------------------------

.RMS <- function(y, n){
  right <- 1 / (n - 1)
  left <- sum(y^2)
  prod <- prod(left, right)
  sqrt(prod)
}
  
.norm_RMS <- function(y, RMS){
  return(y/RMS) # single value
}

norm_RMS <- function(dat, mass_dat, spectra_cols){
  
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  n <- nrow(dat)
  
  rms <- apply(i, 2, FUN = function(x) { .RMS(y = x, n = n)})
  
  i_rms <- t(t(i)*rms^-1)  # i / rms matrix multiplication
  
  dat <- data.frame(mz, i_rms)
  
  colnames(dat) <- c("full_mz", spectra_cols)

  return(dat)
  
}

```

## norm_stdev.R (done)

```{r}
# ------------------------------------
# METHOD: STANDARD DEVIATION OF NOISE
# ------------------------------------

# Find a region of the spectrum with only noise and minimal peaks
# (Should be same region for all spectra)
# Evaluate standard deviation of intensity in that region for each spec
# Divide intensity of EACH PEAK IN SPEC by its st. dev.


norm_stdev <- function(dat, mass_dat, lower = 900, upper= 1100 , spectra_cols){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat)),
    is.numeric(lower),
    is.numeric(upper)
  )
  
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  noise <- i[which(mz > lower & upper > mz),]
  noise[noise == 0] <- NA
  
  std_dev <- apply(noise, 2, FUN = sd, na.rm = TRUE)
  
  i_sd <- t(t(i)*std_dev^-1) # i / st_dev matrix multiplication
      
  dat <- data.frame(mz, i_sd)
  
  colnames(dat) <- c("full_mz", spectra_cols)
  
  return(dat)
}

```

## norm_TIC.R (done)

```{r}
# ----------------------
# METHOD: NORMALIZE TIC
# ----------------------

# This method evaluates the sum of all intensities of each spectrum in a set.
# If the sum is not equivalent between spectra, multiplies the spectra by
# normalization factor.


norm_TIC <- function(dat, mass_dat, spectra_cols){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat)),
    is.numeric(lower),
    is.numeric(upper)
  )    
  
  if(length(spectra_cols) < 2){
    stop("Only one spectrum input. Please enter two spectra for comparison.")
    }
  
  
    mz <- dat[[mass_dat]]
    
    spectra <- lapply(spectra_cols, function(x){dat[x]})
    i <- do.call(what = data.frame, args = c(spectra))

    i_sum <- apply(i, 2, FUN = sum)
  
    max_sum <- max(i_sum)
    
    for(j in 1:length(spectra_cols)){
      
      if(i_sum[j] == max_sum){
      
        other <- i_sum[-j]
        other_index <- which(i_sum %in% other)
        
        factors <- rep(1,length(spectra_cols))
        
        factors[other_index] = i_sum[j]/i_sum[other_index]
        
      }
    }
    
    i_mult <- t(t(i)*factors)
    
    dat <- data.frame(mz, i_mult)
    colnames(dat) <- c("full_mz", spectra_cols)
    
    return(dat)
    
  }

```

## norm_rel_TIC (done)
```{r}
# ---------------------
# METHOD: RELATIVE TIC
# ---------------------

# This method evaluates the sum of all intensities of each spectrum in a set.
# If the sum is not equivalent between spectra, multiplies the spectra by
# normalization factor.
# Each peak in each spectrum is then divided by the normalized TIC.

# Divide intensity of each peak in spectrum by sum of all intensities
.norm_TIC <- function(y, all_y){
  return(y/all_y) # single value
}

norm_rel_TIC <- function(dat, mass_dat, spectra_cols){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat)),
    is.numeric(lower),
    is.numeric(upper)
  )    
  
  
  mz <- dat[[mass_dat]]
    
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))

  i_sum <- apply(i, 2, FUN = sum)
  
  max_sum <- max(i_sum)
  
  for(j in 1:length(spectra_cols)){
      
    if(i_sum[j] == max_sum){
      
      other <- i_sum[-j]
      other_index <- which(i_sum %in% other)
        
      factors <- rep(1,length(spectra_cols))
        
      factors[other_index] = i_sum[j]/i_sum[other_index]
        
      }
    }
  
    i_mult <- t(t(i)*factors)
    
    i_rel_TIC<- t(t(i_mult)*max_sum^-1) # i_mult / max_sum matrix multiplication
    
    dat <- data.frame(mz, i_rel_TIC)
    colnames(dat) <- c("full_mz", spectra_cols)
    
    return(dat)
}

```










## JOSS 6: roxygen2 (Closed)

```{r}
library(Rd2roxygen)

oxygnze <- function(file){
  
  options(roxygen.comment = "##' ")
  str(info <- parse_file(file))
  
  # parse_and_save() combines these two steps
  cat(create_roxygen(info), sep = "\n")
  
}

```



## avgSpectra.Rd (done)
```{r}
oxygnze(file = "/home/sophie/subMaldi/man/avgSpectra.Rd")
```

## baselineCorr.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/baselineCorr.Rd")
```


## createSpecDF.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/createSpecDF.Rd")
```


## find_max.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/find_max.Rd")
```


## find_max_set.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/find_max_set.Rd")
```


## mapSpectrum.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/mapSpectrum.Rd")
```


## normSpectra.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/normSpectra.Rd")
```


## peakDet.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/peakDet.Rd")
```


## readcsvDir.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/readcsvDir.Rd")
```


## readcsvSpec.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/readcsvSpec.Rd")
```


## rmveEmpty.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/rmveEmpty.Rd")
```


## smoothSpectrum.Rd (done)

```{r}
oxygnze(file = "/home/sophie/subMaldi/man/smoothSpectrum.Rd")
```


## subSpectra.Rd (done)
```{r}
oxygnze(file = "/home/sophie/subMaldi/man/subSpectra.Rd")
```




## JOSS 10: test code

## baselineCorr.R

```{r}

```

## baseMALDI.R

```{r}

```

## avgSpectra.R

```{r}

```

## baseline_loess.R

```{r}

```

## baseline_sub.R

```{r}

```

## createSpecDF.R

```{r}

```

## find_max.R

```{r}

```

## find_max_set.R

```{r}

```

## mapSpectrum.R

```{r}

```

## norm_RMS.R

```{r}

```

## norm_TIC.R

```{r}

```

## norm_rel_TIC.R

```{r}

```

## norm_custimp.R

```{r}

```

## norm_custom.R

```{r}

```

## norm_max.R

```{r}

```

## norm_median.R

```{r}

```

## norm_quantile.R

```{r}

```

## norm_stdev.R

```{r}

```

## normSpectra.R

```{r}

library(testthat)

test_that("Method = NULL yields error", {
  expect_error(normSpectra(dat = Master, 
                           mass_dat = "full_mz",
                           method = "apple",
                           norm_mz = NULL,
                           upper = NULL,
                           lower = NULL,
                           spectra_cols = NULL,
                           showHI = FALSE))
})



```

## peakDet.R

```{r}

```

## peak_sub.R

```{r}

```

## plotSpectra.R

```{r}
library(testthat)





```

## readcsvDir.R

```{r}

```

## readcsvSpec.R

```{r}

```

## rmveEmpty.R

```{r}

```

## smoothSpectrum.R

```{r}

```

## smooth_sub.R

```{r}

```

## subSpectra.R

```{r}

```
=======
# normMethod_quantile.R
=======
# normMethod_quantile.R (done)
>>>>>>> d7c9654... Updated issue log

```{r}
# --------------------------------------------------------------------------------------------
# Date: January 28, 2021
# Author: Kristen Yeh, Sophie Castel
# Title: Normalization Method - Quantile
# --------------------------------------------------------------------------------------------


# Quantile normalization consists of two steps:
    # 1. Create a mapping between ranks and values. For rank 1, find the n values,
       # one per spectrum, that are the smallest value on the spectrum, and save their
       # averages. Similarly for rank 2 and the second smallest values, and on up to
       # the n largest values, one per spectrum.
    # 2. For each spectrum, replace the actual values with these averages.

# Basically, index each intensity value in each spectrum then sort
# by intensity (value). Probably need 2 truncate. Average spectrum for the sorted spectra
# is calculated. The averages are then inserted into the sorted spectra.
# The spectra are then reverted to their unsorted order using the index.


.nMethod_quantile <- function(dat, mass_dat, spectra_cols){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat)),
    is.numeric(lower),
    is.numeric(upper)
  )    
  
  if(length(spectra_cols) < 2){
    stop("Only one spectrum input. Please enter two spectra for comparison.")
    }
 
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  i <- data.frame(mz, i)
  colnames(i) <- c("full_mz", spectra_cols)
  
  i_melt <- melt(i, id.vars = "full_mz")
  colnames(i_melt) <- c("full_mz","Spectrum","Intensity")
  
  i_melt <- dplyr::filter(i_melt, Intensity != 0)
  
  i_melt <- i_melt %>% group_by(Spectrum)  %>% mutate(id = row_number())
  ordr <- i_melt[order(i_melt$Intensity, decreasing = TRUE),]
  
  r <- rep(NA, length(spectra_cols))
  
  for(j in 1:length(spectra_cols)){
    r[j] <- max(as.numeric(i_melt$id[which(i_melt$Spectrum == spectra_cols[j])]))
  }
  
  r_min <- min(r)
  
  logic <-which(r == r_min)
  
  trunc <- ordr[(which(ordr$id <= r[logic])),]
  
  i_trunc <- data.frame(matrix(ncol = length(unique(trunc$Spectrum)), nrow = max(trunc$id)))
  
  for(j in 1:length(unique(trunc$Spectrum))){
    
    i_trunc[j] <- trunc$Intensity[which(trunc$Spectrum == spectra_cols[j])]
    colnames(i_trunc)[j] <- spectra_cols[j]
    
  }
  
  i_trunc <- data.frame(i_trunc, avg = apply(i_trunc, 1, FUN = mean, na.rm = TRUE))
  
  for(j in 1:length(unique(trunc$Spectrum))){
    trunc$Intensity[which(trunc$Spectrum == colnames(i_trunc)[j])] <- i_trunc$avg
  }
  
  dat <- trunc[order(trunc[mass_dat], decreasing = FALSE), ] %>% spread(Spectrum, Intensity)
  
  dat[is.na(dat)] <- 0
  dat <- dat[-2]
  return(as.data.frame(dat))
}

```

# normMethod_RMS.R (done)

```{r}

# -------------------------
# METHOD: ROOT MEAN SQUARE
# -------------------------

.RMS <- function(y, n){
  right <- 1 / (n - 1)
  left <- sum(y^2)
  prod <- prod(left, right)
  sqrt(prod)
}
  
.norm_RMS <- function(y, RMS){
  return(y/RMS) # single value
}

.nMethod_RMS <- function(dat, mass_dat, spectra_cols){
  
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  n <- nrow(dat)
  
  rms <- apply(i, 2, FUN = function(x) { .RMS(y = x, n = n)})
  
  i_rms <- t(t(i)*rms^-1)  # i / rms matrix multiplication
  
  dat <- data.frame(mz, i_rms)
  
  colnames(dat) <- c("full_mz", spectra_cols)

  return(dat)
  
}

```

# normMethod_stdev.R (done)

```{r}
# ------------------------------------
# METHOD: STANDARD DEVIATION OF NOISE
# ------------------------------------

# Find a region of the spectrum with only noise and minimal peaks
# (Should be same region for all spectra)
# Evaluate standard deviation of intensity in that region for each spec
# Divide intensity of EACH PEAK IN SPEC by its st. dev.


.normMethod_stdev <- function(dat, mass_dat, lower = 900, upper= 1100 , spectra_cols){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat)),
    is.numeric(lower),
    is.numeric(upper)
  )
  
  mz <- dat[[mass_dat]]
  
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))
  
  noise <- i[which(mz > lower & upper > mz),]
  noise[noise == 0] <- NA
  
  std_dev <- apply(noise, 2, FUN = sd, na.rm = TRUE)
  
  i_sd <- t(t(i)*std_dev^-1) # i / st_dev matrix multiplication
      
  dat <- data.frame(mz, i_sd)
  
  colnames(dat) <- c("full_mz", spectra_cols)
  
  return(dat)
}

```

# normMethod_TIC.R (done)

```{r}
# ----------------------
# METHOD: NORMALIZE TIC
# ----------------------

# This method evaluates the sum of all intensities of each spectrum in a set.
# If the sum is not equivalent between spectra, multiplies the spectra by
# normalization factor.


.normMethod_TIC <- function(dat, mass_dat, spectra_cols){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat)),
    is.numeric(lower),
    is.numeric(upper)
  )    
  
  if(length(spectra_cols) < 2){
    stop("Only one spectrum input. Please enter two spectra for comparison.")
    }
  
  
    mz <- dat[[mass_dat]]
    
    spectra <- lapply(spectra_cols, function(x){dat[x]})
    i <- do.call(what = data.frame, args = c(spectra))

    i_sum <- apply(i, 2, FUN = sum)
  
    max_sum <- max(i_sum)
    
    for(j in 1:length(spectra_cols)){
      
      if(i_sum[j] == max_sum){
      
        other <- i_sum[-j]
        other_index <- which(i_sum %in% other)
        
        factors <- rep(1,length(spectra_cols))
        
        factors[other_index] = i_sum[j]/i_sum[other_index]
        
      }
    }
    
    i_mult <- t(t(i)*factors)
    
    dat <- data.frame(mz, i_mult)
    colnames(dat) <- c("full_mz", spectra_cols)
    
    return(dat)
    
  }

```

# normMethod_rel_TIC (done)

<<<<<<< HEAD
>>>>>>> f435d1f... Updated issue log
=======
```{r}
# ---------------------
# METHOD: RELATIVE TIC
# ---------------------

# This method evaluates the sum of all intensities of each spectrum in a set.
# If the sum is not equivalent between spectra, multiplies the spectra by
# normalization factor.
# Each peak in each spectrum is then divided by the normalized TIC.

# Divide intensity of each peak in spectrum by sum of all intensities
.norm_TIC <- function(y, all_y){
  return(y/all_y) # single value
}

.normMethod_rel_TIC <- function(dat, mass_dat, spectra_cols){
  
  # ---------------------
  # LOGICAL CHECKS
  # ---------------------
  
  stopifnot(
    is.character(mass_dat),
    is.character(spectra_cols),
    mass_dat %in% colnames(dat),
    all(spectra_cols %in% colnames(dat)),
    is.numeric(lower),
    is.numeric(upper)
  )    
  
  
  mz <- dat[[mass_dat]]
    
  spectra <- lapply(spectra_cols, function(x){dat[x]})
  i <- do.call(what = data.frame, args = c(spectra))

  i_sum <- apply(i, 2, FUN = sum)
  
  max_sum <- max(i_sum)
  
  for(j in 1:length(spectra_cols)){
      
    if(i_sum[j] == max_sum){
      
      other <- i_sum[-j]
      other_index <- which(i_sum %in% other)
        
      factors <- rep(1,length(spectra_cols))
        
      factors[other_index] = i_sum[j]/i_sum[other_index]
        
      }
    }
  
    i_mult <- t(t(i)*factors)
    
    i_rel_TIC<- t(t(i_mult)*max_sum^-1) # i_mult / max_sum matrix multiplication
    
    dat <- data.frame(mz, i_rel_TIC)
    colnames(dat) <- c("full_mz", spectra_cols)
    
    return(dat)
}

```
>>>>>>> d7c9654... Updated issue log


